<!-- README.md is generated from README.qmd via: quarto render README.qmd --to gfm -->

# gamlss2: Infrastructure for Flexible Distributional Regression

## Overview 

The primary purpose of this package is to facilitate the creation of advanced infrastructures
designed to enhance the GAMLSS modeling framework. Notably, the `gamlss2` package represents a
significant overhaul of its predecessor, `gamlss`, with a key emphasis on improving estimation
speed and incorporating more flexible infrastructures. These enhancements enable the seamless
integration of various algorithms into GAMLSS, including gradient boosting, Bayesian estimation,
regression trees, and forests, fostering a more versatile and powerful modeling environment.

Moreover, the package expands its compatibility by supporting all model terms from the base
R `mgcv` package. Additionally, the `gamlss2` package introduces the capability to
accommodate more than four parameter families. Essentially, this means that users can now
specify any type of model using these new infrastructures, making the package highly
flexible and accommodating to a wide range of modeling requirements.

* The main model function is `gamlss2()`.
* The default optimizer functions is `RS()`. Optimizer functions can be exchanged.
* Most important methods: `summary()`, `plot()`, `predict()`.
* Easy development of new family objects, see `?family.gamlss2`.
* User-specific "special" terms are possible, see `?special_terms`.

For examples, please visit the manual pages.

```{r help, eval = FALSE}
help(package = "gamlss2")
```

## Installation

The development version of _gamlss2_ can be installed via

```{r installation, eval = FALSE}
install.packages("gamlss2",
  repos = c("https://gamlss-dev.R-universe.dev",
            "https://cloud.R-project.org"))
```

## Licence

The package is available under the
[General Public License version 3](https://www.gnu.org/licenses/gpl-3.0.html)
or [version 2](https://www.gnu.org/licenses/old-licenses/gpl-2.0.html)

## Illustration

```{r preliminaries, echo=FALSE, message=FALSE, results="hide"}
library("gamlss2")
```

The package is designed to follow the workflow of well-established model fitting functions
like `lm()` or `glm()`, i.e., the step of estimating full distributional regression models
is actually not very difficult.

To illustrate the workflow using `gamlss2`, we analyze the `WeatherGermany` data, 

```{r WeatherGermany}
if(!("WeatherGermany" %in% installed.packages())) {
  install.packages('WeatherGermany',
    repos = c("https://gamlss-dev.r-universe.dev",
              "https://cloud.r-project.org"))
}
data("WeatherGermany", package = "WeatherGermany")
head(WeatherGermany)
```

The data consists of daily data and includes the meteorological station `id`s,
the `date` of recording, the maximum wind speed `Wmax` in m/s, the precipitation amount
`pre` in mm, the maximum and minimum temperature `Tmax` (`Tmin`) in degree Celsius, the hours
of sunshine `sun`, the `name` of the station, the `alt`itude in m.a.s.l., as well as the `lon`gitude and
`latìtude coordinates of the station location in Germany.

In this example we use the daily maximum temperature `Tmax` in order estimate a climatology
model using over 37 years of data on the top of Germany at the Zugspitz mountain, the highest
meteorological station in Germany with an altitude of 2956 m.a.s.l.

First, we subset for the Zugspitze

```{r subset}
d <- subset(WeatherGermany, name == "Zugspitze")
```

In order to estimate a good climatology model using _gamlss2_, it is always a good idea to
inspect the distribution of the response

```{r hist_Tmax}
hist(d$Tmax, freq = FALSE, breaks = "Scott")
```

The data seems to be slightly left skewed with longer tales for temperatures below zero. This
is a first hint, that the commonly used normal distribution may not be the most appropriate
distribution for modeling daily maximum temperatures.

Therefore the _gamlss2_ package includes the function `find_family()`, which tries
to identify the most suitable distribution by minimizing an information criterion, the AIC by
default. In this case, we could try out several continuous distributions that are part of the
_gamlss.dist_ package

```{r find_family}
fams <- find_family(d$Tmax,
  families = c(NO, TF, JSU, SEP4))
print(fams)
```

Here, the `SEP4` family seems to have the best fit, we can also visualize the fitted density
with

```{r fit_family}
fit_family(d$Tmax, family = SEP4)
```

After finding a suitable distributional model, we can start including covariates to estimate
a full GAMLSS. Since there is a strong seasonal effect in temperature data, illustrated by
the following scatterplot

```{r season_scatter}
d$yday <- as.POSIXlt(d$date)$yday
plot(Tmax ~ yday, data = d, col = rgb(0.1, 0.1, 0.1, alpha = 0.4))
```

we need to include a model term for seasonal effects. As the _gamlss2_ package supports
all the model terms of the _mgcv_ package we can use the `s()` constructor to set this up.
In addition, we add a time trend to investigate if the maximum temperature has increased
over this period in time. For the full GAMLSS we model each parameter of the selected `SEP4`
distribution. Therefore we first create a new covariate `year` for the overall time trend

```{r time_trend}
d$year <- as.POSIXlt(d$date)$year + 1900
```

afterwards we create the model formula for the four parametric `SEP4` distribution

```{r formula}
f <- Tmax ~ s(yday, bs = "cc") + s(year) |
  s(yday, bs = "cc") + s(year) |
  s(yday, bs = "cc") + s(year) |
  s(yday, bs = "cc") + s(year)
```

Here, the vertical bars `|` seperate the formulas for the parameters of the `SEP4` distribution
and `bs = "cc"` stands for a cyclical spline for the seasonal effect.
Finally, the model is estimated with

```{r model}
b <- gamlss2(f, data = d, family = SEP4)
```

After the model is estimated, we can inspect the model summary output with

```{r summary}
summary(b)
```

Note that the summary output is very similar to `lm()` and `glm()` with the main difference
being that summary outputs are provided for all parameters of the distribution. The summary
shows all estimated linear coefficients (in this case just the intercepts) along the estimated
degrees of freedom for each smooth term in the model. Additionally, the AIC and deviance is reported,
the AIC can also be extracted with

```{r aic}
AIC(b)
```

The log-likelihood using

```{r loglik}
logLik(b)
logLik(b, newdata = d)
```

Here we use the `newdata` argument just to show, that the log-likelihood can also be evaluated on,
e.g., out-of-sample data.

Moreover, estimated effects can be visualized instantly with

```{r effects}
#| fig-height: 6
#| echo: 2
par(mar = c(4, 4, 1, 1))
plot(b, which = "effects")
```

