---
title: "Bayesian GAMLSS with `gamlss2`"
format:
  html:
    html-math-method: mathjax
    toc: true
    number-sections: true
bibliography: gamlss2.bib
vignette: >
  %\VignetteIndexEntry{Bayesian GAMLSS with gamlss2}
  %\VignetteEngine{quarto::html}
  %\VignetteDepends{gamlss2}
  %\VignetteKeywords{Bayesian distributional regression, MCMC}
  %\VignettePackage{gamlss2}
---

Generalized additive models for location, scale, and shape can be estimated by
penalized likelihood or by Bayesian sampling. In `gamlss2`, Bayesian estimation
uses the same distributional formulas, family objects, and smooth terms as the
usual fitting functions. The sampler follows the Bayesian additive model
framework described by @bamlss2018.

There are two convenient ways to run the sampler. Function `bamlss2()` fits a
model directly by MCMC, whereas `mcmc()` starts from an already fitted
`gamlss2` model. This vignette illustrates both workflows and shows how to
inspect posterior effects, samples, and predictions.

## Data and model

We use the `airquality` data and model daily ozone concentration. Ozone is
positive and right-skewed, so a Gamma distribution is a useful starting point.
The location parameter `mu` is modeled as a smooth function of temperature and
wind speed, while the scale parameter `sigma` varies smoothly with
temperature.

```{r}
#| message: false
library("gamlss2")

air <- subset(airquality, !is.na(Ozone))
```

```{r}
#| fig-height: 4.5
#| fig-width: 6
#| fig-align: center
par(mar = c(4, 4, 1, 1))
plot(Ozone ~ Temp, data = air,
  pch = 16, col = adjustcolor(1, 0.3),
  xlab = "Temperature (F)", ylab = "Ozone")
```

The distributional formula is

```{r}
f <- Ozone ~ s(Temp) + s(Wind) | s(Temp)
```

The first right-hand side specifies the predictor for `mu`; the second specifies
the predictor for `sigma`. The same formula can therefore be used for
likelihood-based and Bayesian estimation.

## Bayesian estimation

The convenience function `bamlss2()` sets up the model and runs the Bayesian
sampler. By default, a small number of backfitting iterations is performed
first to obtain starting values before MCMC sampling begins.

```{r}
set.seed(1328)
m <- bamlss2(
  f,
  data = air,
  family = GA,
  n.iter = 1200,
  burnin = 200,
  thin = 1,
  trace = FALSE
)
```

The returned object inherits from class `gamlss2`, so the usual extractor and
plotting methods remain available. In addition, it contains the retained MCMC
samples.

```{r}
summary(m)
```

The sampler uses block Metropolis-Hastings updates for linear and smooth
coefficients. Smooth terms are updated using penalized weighted least squares
proposals, while smoothing variances are sampled by inverse Gamma or slice sampling.
See `?BS` for details of the sampler.

## Posterior effects

Calling `plot()` displays the fitted additive effects. For a Bayesian fit these
are obtained from the posterior samples stored in the fitted object.

```{r}
#| fig-height: 4
#| fig-width: 9
#| fig-align: center
par(mar = c(4, 4, 1, 1))
plot(m)
```

The MCMC draws are available in the `samples` component. For example, the
intercept draws for the distribution parameters can be inspected directly.

```{r}
cn <- grep("(Intercept)", colnames(m$samples),
  fixed = TRUE, value = TRUE)
head(m$samples[, cn, drop = FALSE])
```

## MCMC diagnostics

Posterior summaries are only useful if the sampler has explored the relevant
part of the parameter space adequately. Trace plots and autocorrelation plots
can be obtained with `which = "samples"`. Here we inspect the intercept in the
`mu` model.

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
plot(m, which = "samples", model = "mu",
  terms = "(Intercept)", lag.max = 60)
```

The `model` and `terms` arguments can be used to select other coefficients.
Smooth terms contain several basis coefficients, so inspecting a smooth can
produce several trace and autocorrelation plots.

The values of `n.iter`, `burnin`, and `thin` used above are suitable for a
compact example, but they are not a general convergence rule. For an applied
analysis the chains should be inspected and the number of iterations increased
when necessary.

Sampling can be extended by calling `mcmc()` again on the fitted object. The
existing draws are retained and the newly retained draws are appended to
`samples`, so the call can be repeated as necessary. The `burnin` argument
applies to each additional run.

```{r}
#| eval: false
m <- mcmc(m,
  n.iter = 1200,
  burnin = 200,
  thin = 1,
  trace = FALSE
)
```

## Posterior predictions

For an object returned by `bamlss2()`, `predict()` uses the stored posterior
samples when a summary function is supplied with `FUN`. We evaluate the model
over the observed temperature range and keep wind speed fixed at its median.

```{r}
nd <- data.frame(
  Temp = seq(min(air$Temp), max(air$Temp), length.out = 100),
  Wind = median(air$Wind)
)

post_interval <- function(x) {
  q <- quantile(x, probs = c(0.025, 0.5, 0.975), na.rm = TRUE)
  setNames(q, c("lower", "fit", "upper"))
}

pmu <- predict(m, newdata = nd, model = "mu",
  FUN = post_interval)
psigma <- predict(m, newdata = nd, model = "sigma",
  FUN = post_interval)

head(pmu)
```

The resulting intervals describe posterior uncertainty in the distribution
parameters. They are not prediction intervals for a future ozone observation.

```{r}
#| fig-height: 4
#| fig-width: 9
#| fig-align: center
op <- par(mfrow = c(1, 2), mar = c(4, 4, 1, 1))

plot(Ozone ~ Temp, data = air,
  pch = 16, col = adjustcolor(1, 0.2),
  xlab = "Temperature (F)", ylab = expression(mu))
polygon(c(nd$Temp, rev(nd$Temp)),
  c(pmu$lower, rev(pmu$upper)),
  border = NA, col = adjustcolor(4, 0.15))
lines(nd$Temp, pmu$fit, col = 4, lwd = 2)

plot(nd$Temp, psigma$fit, type = "n",
  ylim = range(psigma[, c("lower", "upper")]),
  xlab = "Temperature (F)", ylab = expression(sigma))
polygon(c(nd$Temp, rev(nd$Temp)),
  c(psigma$lower, rev(psigma$upper)),
  border = NA, col = adjustcolor(4, 0.15))
lines(nd$Temp, psigma$fit, col = 4, lwd = 2)

par(op)
```

Other posterior summaries can be obtained by changing `FUN`. For example,
posterior means and standard deviations are returned with

```{r}
head(predict(m, newdata = nd, type = "parameter", FUN = mean))
head(predict(m, newdata = nd, type = "parameter", FUN = sd))
```

## Starting from a fitted `gamlss2` model

Bayesian sampling can also be started from an existing likelihood-based fit.
This is useful when a model has already been developed with `gamlss2()` and the
same specification should subsequently be sampled. Function `mcmc()` uses the
fitted coefficients and smoothing parameters as starting values and merges the
posterior output back into the model object.

```{r}
m <- gamlss2(f, data = air, family = GA, trace = FALSE)
```

The corresponding MCMC fit is obtained with

```{r}
#| eval: false
set.seed(1328)
m2 <- mcmc(m,
  n.iter = 1200,
  burnin = 200,
  thin = 1,
  trace = FALSE
)
```

Once fitted, `m2` supports the same posterior summaries and plots as the object
returned by `bamlss2()`.

The fitted model supplied to `mcmc()` must retain the design information needed
by the sampler. In particular, reduced model objects that discard this
information cannot be used to restart sampling.

## Practical remarks

The Bayesian and likelihood-based interfaces deliberately share the same model
specification. This makes it possible to develop a distributional model with
`gamlss2()` and switch to MCMC without rewriting its formula or family.
For Bayesian fits, a few points are especially useful:

1. set a seed when reproducible posterior draws are required;
2. inspect trace and autocorrelation plots rather than relying only on posterior
   summaries;
3. increase `n.iter` when the retained draws do not provide adequate mixing;
4. use `predict(..., type = "parameter", FUN = ...)` to summarize uncertainty
   from the stored posterior samples; and
5. distinguish posterior uncertainty in distribution parameters from
   predictive variation in the response.

The general GAMLSS formulation is described in @Rigby+Stasinopoulos:2005. For
the Bayesian additive distributional framework and the MCMC methodology, see
@bamlss2018.
