---
title: "Family Objects"
format:
  html:
    html-math-method: mathjax
    toc: true
    number-sections: true
bibliography: gamlss2.bib
vignette: >
  %\VignetteIndexEntry{Family Objects}
  %\VignetteEngine{quarto::html}
  %\VignetteDepends{gamlss2}
  %\VignetteKeywords{distributional regression, family objects}
  %\VignettePackage{gamlss2}
---

Family objects connect a response distribution with the predictors and
optimization machinery used by `gamlss2`. There are two complementary ways
to define a family:

1. bridge an existing [`distributions3`](https://cran.r-project.org/package=distributions3)
   distribution, or
2. construct a `gamlss2.family` list directly for a distribution that needs
   custom behavior.

The first route should usually be preferred. It gives the fitted model the
same distribution interface used by
`distributions3` for densities,
probabilities, quantiles, moments, and simulation.

```{r preliminaries, echo=FALSE, message=FALSE, results="hide"}
library("gamlss2")
library("distributions3")
```

## The family interface

A `gamlss2.family` object contains the distribution-specific information needed
by the fitting algorithm. The required components are:

| Component | Purpose |
|---|---|
| `family` | Distribution name |
| `names` | Names of the distribution parameters |
| `links` | One link function for each parameter |
| `pdf(par, y, log = FALSE, ...)` | Log-density or density evaluation |

Useful additional components are `cdf()`, `quantile()`, `random()`,
`support()`, `mean()`, `variance()`, `score()`, `hessian()`, `update()`,
`initialize()`, and `valid.response()`. Missing probability, quantile, moment,
and random-number methods can often be constructed numerically from the
density and support, but analytical methods are preferable when they are available.

The two distribution interfaces use slightly different conventions. A family
callback receives a named parameter list, for example `par$mu` and
`par$sigma`, while a `distributions3` method receives an object inheriting from
`distribution`. The bridge described below translates between these forms.

Callback arguments have consistent meanings throughout a family: `par` is a
named list of distribution parameters on the response scale, `y` contains the
observed response values, and `eta` contains linear-predictor values used by an
update method. The `p` argument denotes probabilities for quantiles, `n` is the
requested number of random draws, and `log = TRUE` asks `pdf()` for log-densities.
Additional arguments are passed through `...`.

## Bridging a `distributions3` distribution

The `family()` method turns a `distributions3` object into a
`gamlss2.family`. The names in `links` select the parameters to model and the
values specify their links.

```{r}
fNO <- family(
  Normal(),
  links = c(mu = "identity", sigma = "log")
)

fNO
```

The resulting object contains the family metadata as well as wrappers for
`pdf()`, `cdf()`, `quantile()`, `random()`, the moments, and the log-likelihood.
The distribution constructor is retained in `create_distribution`, which is
used later when a fitted model is converted back to a `distributions3` object.

```{r}
names(fNO)
fNO$names
vapply(fNO$links, function(x) x$name, character(1))
```

The bridge result can now be used like any other family. This example uses the
`cars` data from base R and allows both the normal mean and standard deviation
to vary with speed.

```{r}
m <- gamlss2(
  dist ~ s(speed) | s(speed),
  data = cars,
  family = fNO,
  trace = FALSE
)

summary(m)
```

`prodist()` performs the reverse operation: it extracts the fitted parameters
and returns a vector of `distributions3` objects.

```{r}
nd <- data.frame(speed = c(10, 20, 30))
d <- prodist(m, newdata = nd)
d
```

All ordinary `distributions3` methods can then be used. The `elementwise`
argument controls whether a vector of probabilities is paired with the
distribution vector or expanded to all distribution-probability combinations.

```{r}
mean(d)
variance(d)
cdf(d, 50)
quantile(d, c(0.1, 0.5, 0.9), elementwise = FALSE)
quantile(d, c(0.1, 0.5, 0.9), elementwise = TRUE)
random(d, 2)
```

This gives the complete workflow:

```text
distributions3 object -> family() -> gamlss2.family
        -> gamlss2() -> fitted model -> prodist() -> distributions3 object
```

## Constructor arguments and `distributions3_family()`

The lower-level `distributions3_family()` function is useful when a
distribution constructor has fixed arguments that are not model parameters.
For example, a binomial distribution can have a fixed number of trials while
its probability is modeled.

```{r}
fBI <- distributions3_family(
  "Binomial",
  links = c(p = "logit"),
  create_distribution = "do.call",
  size = 10
)

binomial_distribution <- fBI$create_distribution(
  list(p = rep(0.5, 3))
)
binomial_distribution
distributions3::pdf(binomial_distribution, 1:3)
```

With `create_distribution = "do.call"`, the fixed `size` argument is passed to
the constructor every time a distribution object is created. The default
`"structure"` strategy is faster, but is appropriate only when adding the
class to the parameter object is sufficient.

The workhorse also accepts a custom `create_distribution` function when a
distribution needs a specialized construction step. In all cases, the
returned object must inherit from `gamlss2.family` and its parameter names must
match the supplied links.

## Scores, Hessians, and updates

For a `gamlss2.family` written by hand, `score()` and `hessian()` are functions
on the linear-predictor scale. A `distributions3` score or Hessian is generally
provided with respect to a distribution parameter. The bridge applies the
link-function chain rule when `score = TRUE` or `hessian = TRUE`.

```{r}
fNO <- family(
  Normal(),
  links = c(mu = "identity", sigma = "log"),
  score = TRUE,
  hessian = TRUE,
  update = TRUE
)

names(fNO[c("score", "hessian", "update")])
```

Analytical derivatives can improve speed and numerical stability. If they are
not available, `gamlss2` approximates the required derivatives numerically.
Cross-derivatives can be supplied by the underlying distribution when a
second-order optimizer such as `CG` can benefit from them.

Flexible links are represented by `link-gamlss2` objects. When analytical
score and Hessian callbacks are requested, the bridge uses derivatives from
the link object. For example, the `softplus` link can be passed to the bridge
after construction with `make.link2()`. In this example the derivative flags
are left at their defaults, so numerical derivatives are used.


```{r}
fNO <- family(
  Normal(),
  links = list(mu = make.link2(softplus), sigma = "log")
)

## The softplus link does not provide all second-link derivatives needed by
## the analytic update path, so the bridge uses numerical derivatives here.
m_softplus <- gamlss2(
  dist ~ s(speed) | s(speed),
  data = cars,
  family = fNO,
  trace = FALSE
)

AIC(m, m_softplus)
```

## Numerical completion and support

Families do not need to implement every distributional method. Given a density
and a valid support, `gamlss2` can construct a missing CDF by
integration or summation, invert a CDF to obtain quantiles, calculate moments,
and generate random values by inverse transform sampling.

These fallbacks are convenient for prototypes, but they may be expensive and
need a trustworthy support. In particular, provide `support()` explicitly for
shifted or truncated distributions, unusual continuous supports, and discrete
families that are not counts on the nonnegative integers. Analytical CDFs and
quantiles should be retained whenever possible.

The `type` component should be either `"continuous"` or `"discrete"`, and
`valid.response()` should reject invalid response values. The type and support
are used by residual diagnostics, numerical completion, and the
`distributions3` methods.

## Standard and user-defined families

The package is being extended with distributions3-backed versions of the
standard `gamlss.dist` families. These will use the familiar `f*` names, such
as `fNO`, `fGA`, `fBI`, and `fNBI`, while exposing the same probability,
quantile, moment, simulation, and derivative infrastructure. For these standard
families, users should normally select the supplied family rather than rewrite
the distribution callbacks themselves.

A user-defined family is still useful for a new distribution, a modified
parameterization, or specialized initialization and updating. The next section
shows the minimal pattern; `?gamlss2.family` documents the family components,
and `?family.distribution` documents the distributions3 bridge. In particular,
the `score`, `hessian`, and optional `update` callbacks are the extension points
for custom fitting behavior.

## Writing a family directly

When no suitable `distributions3` distribution exists, a family can be written
as a classed list. The following is a compact normal family with analytical
probability and moment methods. The `random()` signature is
`random(par, n)`, matching the `gamlss2.family` interface.

```{r}
my_normal_family <- function(mu.link = "identity", sigma.link = "log") {
  fam <- list(
    family = "my_normal_family",
    names = c("mu", "sigma"),
    links = list(mu = mu.link, sigma = sigma.link),
    pdf = function(par, y, log = FALSE, ...) {
      dnorm(y, mean = par$mu, sd = par$sigma, log = log)
    },
    cdf = function(par, y, ...) {
      pnorm(y, mean = par$mu, sd = par$sigma, ...)
    },
    quantile = function(par, p, ...) {
      qnorm(p, mean = par$mu, sd = par$sigma)
    },
    random = function(par, n, ...) {
      rnorm(n, mean = par$mu, sd = par$sigma)
    },
    mean = function(par, ...) par$mu,
    variance = function(par, ...) par$sigma^2,
    support = c(-Inf, Inf),
    type = "continuous",
    valid.response = function(y) is.numeric(y)
  )
  class(fam) <- "gamlss2.family"
  fam
}

my_family <- my_normal_family()

m <- gamlss2(
  dist ~ s(speed) | s(speed),
  data = cars,
  family = my_family,
  trace = FALSE
)

summary(m)
```

For a manually defined family, the minimum required interface is `family`,
`names`, `links`, and `pdf()`. Derivative callbacks have the following shape:

```r
score = list(
  mu = function(par, y, ...) {
    (y - par$mu) / par$sigma^2
  },
  sigma = function(par, y, ...) {
    -1 + (y - par$mu)^2 / par$sigma^2
  }
)

hessian = list(
  mu = function(par, y, ...) 1 / par$sigma^2,
  sigma = function(par, y, ...) rep(2, length(y)),
  "mu:sigma" = function(par, y, ...) rep(0, length(y))
)
```

The Hessian entries are the negative of the expected second derivatives on
the predictor scale.
An optional `update(par, y, eta, which)` function can provide a specialized
Newton/Fisher-scoring update returning `list(eta, weights)`.

## Optional scoring rules

The distributions3 bridge also exposes a `crps()` callback when the optional
`scoringRules` package is available. This is useful for probabilistic forecast
evaluation, but it is not needed for fitting, prediction, or the basic family

## Summary

Use the `distributions3` bridge when the response distribution already has a
`distributions3` implementation. It provides a consistent probability and
moment interface and handles link-scale derivative transformations. Use a
manual `gamlss2.family` when a distribution needs custom density, derivative,
support, initialization, or optimization logic. In either case, a good family
should provide reliable support information and analytical distributional
methods whenever practical.

