---
title: "Family Extensions"
format:
  html:
    html-math-method: mathjax
    toc: true
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Family Extensions}
  %\VignetteEngine{quarto::html}
  %\VignetteDepends{gamlss2}
  %\VignetteKeywords{distributional regression, family objects, truncation, censoring, inflation}
  %\VignettePackage{gamlss2}
---

A response distribution should reflect how the observations were obtained.
An earthquake sample may exclude small events, a measuring device may stop
at a fixed limit, and a proportion may include exact zeros and ones. In each
case, a familiar distribution can be a useful starting point, but its support
or probabilities need to change before it describes the recorded data.

Family extensions make these changes to an existing family. The result
is another `gamlss2.family`, ready to use in `fit_family()` or `gamlss2()`.
We begin with earthquake magnitudes and use them to distinguish selection
from rounding and grouping. We then consider recording limits, extra point
masses, and transformations. Each example connects the
observation mechanism to a fitted model.

```{r preliminaries}
#| message: false
#| results: hide
library("gamlss2")
ctrl <- gamlss2_control(trace = FALSE, maxit = 100)
```

## Choosing an observation model

A family extension specifies how a base distribution becomes the
the response we observe. Several extensions can be combined when more than
one mechanism is involved.

For example, earthquake magnitude can be viewed as a latent continuous
quantity. If only events above a threshold are retained and magnitude is
reported to one decimal place, its observed family is obtained by truncating
the continuous family first and rounding second.

The available constructors cover the following operations.

| Constructor | Operation | Resulting parameters |
|---|---|---|
| `transform_family()` | Fixed monotone transformation | Base parameters |
| `logit_family()` | Logistic transformation to $(0, 1)$ | Base parameters |
| `truncate_family()` | Condition on a fixed interval | Base parameters |
| `censor_family()` | Fixed censoring limits or a right-censored `Surv()` response | Base parameters |
| `inflate_family()` | Add one or more point masses | Base parameters plus `pi1`, `pi2`, ... |
| `hurdle_family()` | Add one point mass and exclude it from the base component | Base parameters plus `pi1` |
| `discretize_family()` | Assign latent intervals to fixed values | Base parameters |
| `round_family()` | Round a continuous response to fixed decimal precision | Base parameters |

The constants defining the operation---bounds, atom locations, bin breaks,
and transformations---are fixed when the family is constructed. They are not
additional distributional parameters. Except for inflation and hurdle odds,
the parameter names, order, and links of the base family are retained.

## Truncation: earthquake magnitudes

The base R `quakes` data contain 1,000 seismic events near Fiji. According to
the data description, only events with body-wave magnitude greater than 4.0
were retained. Before choosing a model, it is useful to inspect the response.
Its observed minimum is exactly 4.0, and its strongly asymmetric shape at that
boundary is immediately visible in a histogram.

```{r quakes-data}
data("quakes", package = "datasets")

range(quakes$mag)
min(quakes$mag)
```

The histogram shows the lower boundary and a long right tail. The scatterplot
suggests that magnitude also varies with depth. We first investigate the
boundary; later we return to the one-decimal reporting precision.

```{r quakes-plot}
#| fig-height: 4.2
#| fig-width: 8
#| fig-align: center
op <- par(mfrow = c(1, 2), mar = c(4, 4, 1, 1))
hist(quakes$mag,
  breaks = seq(3.95, max(quakes$mag) + 0.05, by = 0.1),
  right = FALSE, border = "white", col = "grey65",
  xlab = "Magnitude", main = "")
plot(mag ~ depth, data = quakes,
  pch = 16, col = adjustcolor(1, 0.2),
  xlab = "Depth (km)", ylab = "Recorded magnitude")
par(op)
```

We first treat the recorded values as numeric measurements and compare an
ordinary normal working model with a normal distribution truncated at 4. The
endpoint is included so that the recorded value 4.0 is valid; for a continuous
latent distribution this does not alter the normalizing probability.

```{r quakes-families}
f_norm <- NO
f_trunc <- truncate_family(
  NO, lower = 4, include = c(TRUE, TRUE)
)
```

Truncation divides the original density by the probability of falling in the
retained interval. For a continuous distribution with a lower boundary at 4,
this gives $f(y \mid Y \geq 4) = f(y)/\{1-F(4)\}$. The parent normal location
may lie below 4; it is not the mean of the observed magnitudes. This is how a
symmetric parent distribution can produce an asymmetric observed response.

Before adding covariates, we check whether this change helps describe the
marginal distribution. `fit_family()` fits an intercept-only `gamlss2()` model
and overlays its density on a histogram.

```{r quakes-marginal}
#| fig-width: 8
#| fig-height: 4
op <- par(mfrow = c(1, 2))
mf_norm <- fit_family(quakes$mag, family = NO,
  control = ctrl, main = "Normal", xlab = "Magnitude")
mf_trunc <- fit_family(quakes$mag,
  family = f_trunc,
  control = ctrl, main = "Truncated normal", xlab = "Magnitude")
par(op)
AIC(mf_norm, mf_trunc)
```

The truncated normal has substantially lower AIC. Compare the lower boundary
and the right tail in the two plots to see where the distributions differ.
The AIC improvement supports this choice, but does not establish the sampling
mechanism: the
known selection threshold motivates truncation. Both fits use continuous
densities for the same response, so their AIC values are comparable. Treating
the recorded decimals as continuous is a working approximation.

The marginal comparison also ignores depth and geographic location. We next
use `s(depth)` and a spatial smooth `s(long, lat, k = 30)`, keeping scale
constant. Applying the same regression specification to both families checks
whether truncation remains useful after accounting for these covariates.

```{r quakes-fit}
m_norm <- gamlss2(
  mag ~ s(depth) + s(long, lat, k = 30) | 1,
  data = quakes,
  family = f_norm,
  control = ctrl
)

m_trunc <- gamlss2(
  mag ~ s(depth) + s(long, lat, k = 30) | 1,
  data = quakes,
  family = f_trunc,
  control = ctrl
)

AIC(m_norm, m_trunc)
```

The truncated family also has lower AIC after including the smooth depth
and spatial effects. Thus its marginal advantage was not simply a consequence
of pooling events from different depths and locations.

## Rounding: reported measurement precision

The lower threshold explains which events enter the sample. Reporting
precision explains why magnitudes occur on a one-decimal grid. We can include
both features in the same model. `round_family()`
changes the likelihood contribution from a density at the recorded number to
the probability of the latent rounding interval. The response column remains
numeric in R, but its statistical treatment is now discrete. For example,
a reported value of 4.5 represents an interval approximately from 4.45 to
4.55. Its probability is a difference of CDF values over that interval.
With truncation at 4 followed by rounding, the reported value 4.0 represents
only the retained part of its rounding interval, from 4 to 4.05.

```{r quakes-rounded-fit}
f_round <- round_family(NO, digits = 1)
f_both <- round_family(
  truncate_family(NO, lower = 4),
  digits = 1
)

m_round <- gamlss2(
  mag ~ s(depth) + s(long, lat, k = 30) | 1,
  data = quakes,
  family = f_round,
  control = ctrl
)

m_both <- gamlss2(
  mag ~ s(depth) + s(long, lat, k = 30) | 1,
  data = quakes,
  family = f_both,
  control = ctrl
)

AIC(m_round, m_both)
```

This second AIC comparison is valid because both models assign probability
mass to the same one-decimal values. In contrast, neither rounded model should be compared by log-likelihood or
AIC with `m_norm` or `m_trunc`:
probability densities and probability masses are defined with respect to
different measures and have different units.

Parameters returned by `predict(..., type = "parameter")` still describe the
parent normal family, before selection and rounding. In particular, `mu` is
not the mean of the observed, truncated distribution. Distributional methods
operate on the final recorded scale. Thus, the following curves are rounded
predictive quantiles from the truncated-and-rounded model.

```{r quakes-prediction}
nd <- data.frame(
  depth = seq(min(quakes$depth), max(quakes$depth), length.out = 150),
  long = median(quakes$long),
  lat = median(quakes$lat)
)
q <- quantile(m_both, newdata = nd,
  probs = c(0.1, 0.5, 0.9))

head(predict(m_both, newdata = nd, type = "parameter"))
```

These curves hold longitude and latitude at their sample medians. The points
include all locations, so they show the overall data cloud rather than a
sample drawn at that fixed location.

```{r quakes-quantiles}
#| fig-height: 5
#| fig-width: 7
#| fig-align: center
par(mar = c(4, 4, 1, 1))
plot(mag ~ depth, data = quakes,
  pch = 16, col = adjustcolor(1, 0.15),
  xlab = "Depth (km)", ylab = "Recorded magnitude")
matlines(nd$depth, q,
  lty = c(2, 1, 2), lwd = 2, col = 4)
legend("topright", c("10%", "50%", "90%"),
  lty = c(2, 1, 2), lwd = 2, col = 4, bty = "n")
```

Order is part of the model. The family used above selects a latent event and
then rounds its magnitude. Rounding first and then truncating would instead
select observations according to the displayed value, which is a different
sampling mechanism near the boundary.

## Discretization: measurements in groups

Rounding uses equal-width intervals determined by decimal precision.
`discretize_family()` generalizes this idea to arbitrary, fixed bins. Each
reported value identifies an interval, and the likelihood uses the probability
of that interval rather than a density at its representative value.

To illustrate, suppose the earthquake depths were released only in four
groups. We create this coarsened response from `quakes`; these groups are an
illustration, not how the original data were recorded. With `right = TRUE`,
the intervals are right-closed:

```{r discretized-family}
breaks <- c(-Inf, 100, 300, 500, Inf)
values <- c(50, 200, 400, 600)

f_group <- discretize_family(
  NO,
  breaks = breaks,
  values = values,
  right = TRUE
)

quakes$depth_released <- values[
  cut(quakes$depth, breaks = breaks, labels = FALSE, right = TRUE)
]
table(quakes$depth_released)
```

Each mass is an interval probability under the latent family, rather than a
density evaluated at the representative value. We can model the latent normal
location as a function of magnitude while fitting the four released values
directly.

```{r discretized-fit}
m_group <- gamlss2(
  depth_released ~ s(mag) | 1,
  data = quakes,
  family = f_group,
  control = ctrl
)

head(predict(m_group, type = "parameter"))
quantile(m_group,
  newdata = data.frame(mag = c(4, 5, 6)),
  probs = c(0.1, 0.5, 0.9)
)
```

`round_family()` is the specialized version for a regular decimal grid. It
uses intervals centered on the displayed values and follows the convention of
base R's `round()`. Both constructors return discrete families and therefore
use randomized quantile residuals when fitted.

## Censoring: measurements at a recording limit

Truncation excludes events from the sample; rounding and grouping record
intervals. A recording limit
has a different consequence: the observation remains in the data, but we
know only that its true value reached or crossed the limit. This is censoring.
For upper censoring at $u$, every $X \geq u$ is observed as $u$;
the resulting distribution has a point mass of size $\Pr(X \geq u)$ at the
boundary.

The distinction can be illustrated by artificially capping the base R `cars`
stopping distances at 80 feet. The pile-up at 80 is evidence of censoring, not
truncation.

```{r censoring-data}
data("cars", package = "datasets")
cars$dist80 <- pmin(cars$dist, 80)

table(cars$dist80 == 80)
```

```{r censoring-plot}
#| fig-height: 5
#| fig-width: 7
#| fig-align: center
par(mar = c(4, 4, 1, 1))
plot(dist80 ~ speed, data = cars,
  pch = 16, col = adjustcolor(1, 0.55),
  xlab = "Speed (mph)", ylab = "Recorded stopping distance (ft)")
abline(h = 80, lty = 2, col = 2)
```

A censored normal model assigns the full upper-tail probability to every
recorded value at 80. It can be fitted directly with `gamlss2()`:

```{r censoring-fit}
f_cens <- censor_family(NO, upper = 80)

# Use the ordinary normal fit only to obtain stable starting values.
m_start <- gamlss2(
  dist80 ~ s(speed) | .,
  data = cars,
  family = NO,
  control = ctrl
)

m_cens <- gamlss2(
  dist80 ~ s(speed) | .,
  data = cars,
  family = f_cens,
  control = ctrl,
  start = coef(m_start)
)

head(predict(m_cens, type = "parameter"))
```

The fitted quantiles show how the conditional distribution changes with
speed. Because the recorded response cannot exceed 80, upper quantiles flatten
at the censoring limit when enough fitted probability lies above it.

```{r censoring-quantiles}
#| fig-height: 5
#| fig-width: 7
nd <- data.frame(speed = seq(min(cars$speed), max(cars$speed), length.out = 150))
q <- quantile(m_cens, newdata = nd, probs = c(0.1, 0.5, 0.9))
par(mar = c(4, 4, 1, 1))
plot(dist80 ~ speed, data = cars,
  pch = 16, col = adjustcolor(1, 0.45),
  xlab = "Speed (mph)", ylab = "Recorded stopping distance (ft)",
  ylim = range(cars$dist80, q))
matlines(nd$speed, q, lty = c(2, 1, 2), lwd = 2, col = 4)
abline(h = 80, lty = 3, col = 2)
legend("left", c("10%", "50%", "90%", "Limit"),
  lty = c(2, 1, 2, 3), lwd = c(2, 2, 2, 1),
  col = c(4, 4, 4, 2), bty = "n")
```

For this example, the chance of reaching the cap is an interpretable fitted
quantity. It is the normal upper-tail probability under the predicted parent
parameters, and also the point mass assigned to the recorded value 80.

```{r censoring-prediction}
at <- data.frame(speed = c(10, 15, 20, 25))
d <- predict(m_cens, newdata = at)
data.frame(at,
  probability_at_limit = 1 - cdf(d, 80))
```

Unlike censoring, upper truncation would remove cars whose stopping distance
exceeds 80 and renormalize the density among those that remain. For two-sided
mechanisms, supply both `lower` and `upper`. `truncate_family()` additionally
accepts `include`, which controls inclusion of each endpoint and matters for
discrete base families.

## Right-censored survival responses

The fixed boundaries of `censor_family()` describe a recording device that
caps every response at the same known value. Survival data are different:
each individual can be right-censored at a different observed time, and the
response therefore includes both time and event status.

The `lung` data from the `survival` package illustrate the appropriate
interface. Status code 2 denotes an observed death; code 1 denotes
right-censoring.

```{r survival-data}
lung <- survival::lung
lung$event <- lung$status == 2
lung$sex <- factor(lung$sex, levels = 1:2,
  labels = c("Male", "Female"))

table(lung$event)
```

Calling `censor_family(WEI)` without a fixed boundary constructs a family for
observation-specific right censoring. It accepts a `Surv()` response directly:
an event contributes the Weibull density at its observed time, whereas a
right-censored observation contributes the probability of surviving beyond
that time. In the `WEI` parameterization, `mu` is the Weibull scale and
`sigma` its shape. We model scale by age and sex and keep shape constant.

```{r survival-fit}
f_surv <- censor_family(WEI)

m_surv <- gamlss2(
  survival::Surv(time, event) ~ s(age) + sex | 1,
  data = lung,
  family = f_surv,
  control = ctrl
)

logLik(m_surv)
head(predict(m_surv, type = "parameter"))
```

The likelihood contribution is a density when `event` is true and a survival
probability when it is false. Thus the censoring information remains attached
to every observation instead of being encoded by replacing time with one
fixed boundary value.

To see the covariate effects on the response scale, predict survival curves
for ages 50, 65, and 80, separately for men and women. Predictions use the
underlying event-time distribution: $S(t \mid x) = 1-F(t \mid x)$.
The censoring indicator is needed for fitting, but is not a predictor of
survival. These are model-based conditional associations.

```{r survival-prediction}
nd <- expand.grid(age = c(50, 65, 80), sex = c("Male", "Female"))
d <- predict(m_surv, newdata = nd)
time <- seq(0, max(lung$time), length.out = 200)
surv <- sapply(seq_len(nrow(nd)), function(j) {
  1 - cdf(d[j], time)
})
```

```{r survival-plot}
#| fig-width: 8
#| fig-height: 4.5
op <- par(mfrow = c(1, 2), mar = c(4, 4, 4, 1))
for(i in c("Male", "Female")) {
  j <- which(nd$sex == i)
  matplot(time, surv[, j], type = "l", lty = 1:3,
    col = c("#0072B2", "#D55E00", "#009E73"), lwd = 2,
    ylim = c(0, 1), xlab = "Time (days)", ylab = "Survival probability",
    main = i)
  legend("topright", paste("Age", nd$age[j]), lty = 1:3,
    col = c("#0072B2", "#D55E00", "#009E73"), lwd = 2, bty = "n")
}
par(op)
```

## Inflation and hurdles: two ways to model zeros

Inflation adds a source of zeros to an ordinary count distribution. A hurdle
model instead models zero versus positive separately, then uses the base
distribution conditional on being positive. If $p$ is the added-component
probability and the Poisson mean is $\mu$, the respective zero probabilities
are

$$
\Pr_{\mathrm{inflated}}(Y=0)=p+(1-p)e^{-\mu},
\qquad
\Pr_{\mathrm{hurdle}}(Y=0)=p.
$$

Thus a hurdle can describe either excess zeros or fewer zeros than the
Poisson distribution predicts. Inflation can only add zeros. In both
constructors, `pi1` is the odds $p/(1-p)$, with a log link; it is not the
overall zero probability in an inflated model.

A further distinction matters when comparing fitted models: for a constant
Poisson mean and an excess of zeros, the two models can represent exactly the
same count distribution after changing the zero-component probability.
Matching fitted curves are therefore possible even when the mechanisms have
different interpretations. Counts alone do not identify which mechanism
produced those zeros.

### Checking the distributions before regression

We simulate two samples to make these properties visible without confounding
them with a poorly fitting positive tail. Both use a Poisson component with
mean 1.5. In the first sample, an additional zero component has probability
0.35. In the second, a hurdle assigns only 0.05 probability to zero, below
the ordinary Poisson probability $e^{-1.5}$ of about 0.22.

```{r count-data}
set.seed(731)
f_infl <- inflate_family(PO, at = 0)
f_hurd <- hurdle_family(PO, at = 0)
ys <- list(
  "Excess zeros" = f_infl$random(
    list(mu = 1.5, pi1 = 0.35 / 0.65), 800),
  "Few zeros" = f_hurd$random(
    list(mu = 1.5, pi1 = 0.05 / 0.95), 800)
)
vapply(ys, function(y) mean(y == 0), numeric(1))
```

Fit all three distributions to each sample. The plots compare empirical
frequencies with fitted probabilities over the same count range.

```{r count-marginal}
#| fig-width: 9
#| fig-height: 6
op <- par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))
fits <- lapply(names(ys), function(label) {
  y <- ys[[label]]
  fits <- list(
    Poisson = fit_family(y, family = PO, control = ctrl,
      main = paste(label, "Poisson", sep = ": "), xlab = "Count",
      xlim = c(0, 8), ylim = c(0, 0.65)),
    Inflated = fit_family(y, family = f_infl, control = ctrl,
      main = paste(label, "Inflated", sep = ": "), xlab = "Count",
      xlim = c(0, 8), ylim = c(0, 0.65)),
    Hurdle = fit_family(y, family = f_hurd, control = ctrl,
      main = paste(label, "Hurdle", sep = ": "), xlab = "Count",
      xlim = c(0, 8), ylim = c(0, 0.65))
  )
  fits
})
par(op)
names(fits) <- names(ys)
lapply(fits, function(fits) AIC(fits$Poisson, fits$Inflated, fits$Hurdle))
```

For excess zeros, inflation and a hurdle should both follow the data, and
can give essentially identical fits. For the second sample, the hurdle can
represent the low zero probability while preserving the positive-count
shape. The inflated fit approaches ordinary Poisson as its extra-zero odds
approach zero, but cannot remove natural Poisson zeros. AIC comparisons here
are within each sample, never between the two different samples.

### Count regression

We next simulate a regression example in which both the positive counts and
the probability of zero change with `x`. This lets us check whether the
fitted model recovers each part of the distribution.

```{r count-regression}
set.seed(732)
dat <- data.frame(x = runif(1000, -2, 2))
dat$y <- f_hurd$random(list(
  mu = exp(0.3 + 0.5 * sin(dat$x)),
  pi1 = exp(-1.5 + 0.8 * dat$x)
), nrow(dat))

m_hurd <- gamlss2(y ~ s(x) | s(x), data = dat,
  family = f_hurd, control = ctrl)
nd <- data.frame(x = seq(-2, 2, length.out = 150))
p <- predict(m_hurd, newdata = nd, type = "parameter")
p0 <- p$pi1 / (1 + p$pi1)
mpos <- p$mu / (-expm1(-p$mu))
```

Here `mu` is the parent Poisson mean, not the mean of the positive counts.
Conditioning on positivity gives $E(Y \mid Y>0,x)=\mu/(1-e^{-\mu})$.
The plots compare these fitted quantities with the known simulation curves.

```{r count-regression-plot}
#| fig-width: 8
#| fig-height: 4
op <- par(mfrow = c(1, 2), mar = c(4, 4, 1, 1))
plot(nd$x, p0, type = "l", col = 4, lwd = 2,
  ylim = c(0, 1), xlab = "x", ylab = "Probability of zero")
lines(nd$x, plogis(-1.5 + 0.8 * nd$x), lty = 2, lwd = 2)
legend("topleft", c("Fitted", "Generating model"),
  col = c(4, 1), lty = c(1, 2), lwd = 2, bty = "n")
mu <- exp(0.3 + 0.5 * sin(nd$x))
mpos0 <- mu / (-expm1(-mu))
plot(nd$x, mpos, type = "l", col = 4, lwd = 2,
  ylim = range(c(mpos, mpos0)),
  xlab = "x", ylab = "Mean among positive counts")
lines(nd$x, mpos0, lty = 2, lwd = 2)
par(op)
```

### Proportions with exact endpoints

Point masses are also useful for continuous responses. Several inflation
points are allowed, giving a model for proportions that can attain both
endpoints while otherwise varying continuously. `logit_family(NO)` maps a normal variable through the logistic
function, producing values strictly between 0 and 1. Inflation then adds the
two endpoints. Here we simulate data with a known mechanism so that the role
of each component is clear. `mu` and `sigma` govern the latent
logistic-normal component, while `pi1` and `pi2` are odds for exact zero and
exact one, respectively.

```{r endpoint-fit}
set.seed(2026)
dat <- data.frame(x = runif(250, -2, 2))
f_01 <- inflate_family(
  logit_family(NO),
  at = c(0, 1)
)
dat$proportion <- f_01$random(
  list(
    mu = sin(1.5 * dat$x),
    sigma = exp(-1 + cos(dat$x)),
    pi1 = 0.15,
    pi2 = 0.10
  ),
  nrow(dat)
)

m_01 <- gamlss2(
  proportion ~ s(x) | s(x),
  data = dat,
  family = f_01,
  control = ctrl
)

head(predict(m_01, type = "parameter"))
```

We can now inspect fitted quantiles across `x`. Because these describe the
complete distribution, they include the endpoint masses as well as the
continuous interior.

```{r endpoint-plot}
#| fig-width: 7
#| fig-height: 5
nd <- data.frame(x = seq(-2, 2, length.out = 150))
q <- quantile(m_01, newdata = nd,
  probs = c(0.1, 0.5, 0.9))
par(mar = c(4, 4, 1, 1))
plot(proportion ~ x, data = dat,
  pch = 16, col = adjustcolor(1, 0.3), ylab = "Proportion")
matlines(nd$x, q,
  lty = c(2, 1, 2), lwd = 2, col = 4)
legend("topleft", c("10%", "50%", "90%"),
  lty = c(2, 1, 2), lwd = 2, col = 4, bty = "n")
```

The smooth estimates how the interior distribution changes with `x`; the
endpoint odds are constant in this example. The four right-hand sides
correspond, in order, to `mu`, `sigma`, `pi1`, and
`pi2`. Exact endpoints are handled as masses; interior observations retain a
continuous density contribution.

## Transformations: modeling a positive response

`transform_family()` describes the distribution of $Y = g(X)$ when $X$
follows a continuous base family and $g$ is a fixed monotone transformation. In addition to `g`, it needs
the inverse transformation and the log absolute Jacobian of the inverse:

$$
\log f_Y(y) = \log f_X\{g^{-1}(y)\}
  + \log \left|\frac{d g^{-1}(y)}{dy}\right|.
$$

The logistic transformation above restricts a response to an interval.
Exponentiating a normal variable instead produces a positive, right-skewed
lognormal distribution. Tree volume in the base R `trees` data provides an
example. The Jacobian ensures that the transformed density integrates to
one and that the likelihood is expressed on the original volume scale.

```{r transformed-family}
f_lnorm <- transform_family(
  family = NO,
  transform = exp,
  inverse = log,
  log_jacobian = function(y) -log(y),
  support = c(0, Inf),
  valid.response = function(y) {
    is.na(y) | (is.finite(y) & y > 0)
  },
  name = "Lognormal"
)

data("trees", package = "datasets")
```

First compare normal and lognormal marginal fits. Both are densities on the
volume scale, so their AIC values are comparable. A model fitted to
`log(Volume)` without the Jacobian would use a different response scale.

```{r transformed-marginal}
#| fig-width: 8
#| fig-height: 4
op <- par(mfrow = c(1, 2))
mf_norm <- fit_family(trees$Volume, family = NO,
  control = ctrl, main = "Normal", xlab = "Volume")
mf_lnorm <- fit_family(trees$Volume, family = f_lnorm,
  control = ctrl, main = "Lognormal", xlab = "Volume")
par(op)
AIC(mf_norm, mf_lnorm)
```

The lognormal has lower marginal AIC in this comparison. This is a useful
starting point, but the marginal distribution combines trees of different
sizes and does not determine the best conditional family. We now model the
location of log volume as a smooth function of girth and inspect conditional
quantiles on the original volume scale.

```{r transformed-fit}
m_lnorm <- gamlss2(
  Volume ~ s(Girth) | 1,
  data = trees,
  family = f_lnorm,
  control = ctrl
)
```

```{r transformed-prediction}
nd <- data.frame(Girth = seq(min(trees$Girth), max(trees$Girth),
  length.out = 100))
q <- quantile(m_lnorm, newdata = nd,
  probs = c(0.1, 0.5, 0.9))
```

```{r transformed-plot}
#| fig-height: 5
#| fig-width: 7
#| fig-align: center
par(mar = c(4, 4, 1, 1))
plot(Volume ~ Girth, data = trees,
  pch = 16, col = adjustcolor(1, 0.55),
  xlab = "Girth (in)", ylab = "Volume (ft^3)",
  ylim = range(trees$Volume, q))
matlines(nd$Girth, q,
  lty = c(2, 1, 2), lwd = 2, col = 4)
legend("topleft", c("10%", "50%", "90%"),
  lty = c(2, 1, 2), lwd = 2, col = 4, bty = "n")
```

Decreasing transformations are supported with `increasing = FALSE`.
`logit_family()` is the ready-made logistic transformation of a continuous
family. It has strict support $0 < Y < 1$; exact endpoints require a different
mechanism, such as the inflated endpoint family above.

## Practical guidance

The family extension should describe how a latent response becomes an observed
response. In practice:

1. Start with the family for the latent response.
2. Apply extensions from the inside out in the order the mechanisms occur.
3. Use `fit_family()` to inspect marginal fit. Compare AIC only for models
   fitted to the same response with compatible likelihoods.
4. Add covariates with `gamlss2()`, using `s()` for nonlinear effects, and
   check whether the extended family remains useful conditionally.
5. Interpret `predict(..., type = "parameter")` on the base-family scale, but
   interpret probabilities, quantiles, simulations, and residuals under the
   fully extended family.

A few boundaries of the infrastructure are intentional. Constants passed to
the extension constructors are fixed. `censor_family()` also accepts
observation-specific
right-censoring times through a `Surv()` response; other censoring types need
a specialized family. Estimated transformation parameters require a family
that explicitly includes them. Generic censored and rounded means and
variances are not reported because exact calculation would require
partial moments that are not part of the family interface. Arbitrary weighted
distributions, splices, non-monotone transformations, convolutions, and
compound distributions are likewise outside this infrastructure.

Finally, match the family to the actual observation process. A pile-up at a
detection limit suggests censoring; missing out-of-range cases imply
truncation; extra and natural zeros together imply inflation; and a separate
zero-generating process suggests a hurdle. These choices are scientifically
different even when the observed histograms look similar.

See `?family_extensions` for the complete argument and return-value
reference for all constructors.
