---
title: "First Steps"
format: 
  html:
    html-math-method: mathjax
    toc: true
    number-sections: true
bibliography: gamlss2.bib
vignette: >
  %\VignetteIndexEntry{First Steps}
  %\VignetteEngine{quarto::html}
  %\VignetteDepends{gamlss2}
  %\VignetteKeywords{distributional regression}
  %\VignettePackage{gamlss2}
---

Generalized additive models for location, scale, and shape (GAMLSS) model
all parameters of a response distribution rather than only its mean
[@Rigby+Stasinopoulos:2005]. The `gamlss2` package provides a workflow similar
to `lm()` and `glm()`, while allowing a separate additive predictor for every
distribution parameter.

This vignette illustrates the basic workflow:

1. specify and fit a distributional regression model,
2. inspect its estimated effects and residuals, and
3. obtain distribution parameters, probabilities, and quantiles.

## Data

We use the `HarzTraffic` data included in the package. They contain daily
traffic counts and weather measurements near Sonnenberg in the Harz region of
Germany. Motorcycle traffic has a pronounced annual cycle, so we model the
number of motorcycles (`bikes`) as a smooth function of the day of the year
(`yday`).

```{r}
#| message: false
library("gamlss2")
data("HarzTraffic", package = "gamlss2")
```

```{r}
#| fig-height: 4.5
#| fig-width: 6
#| fig-align: center
par(mar = c(4, 4, 1, 1))
plot(bikes ~ I(yday + 1), data = HarzTraffic,
  pch = 16, col = adjustcolor(1, 0.25),
  xlab = "Day of year", ylab = "Motorcycles")
```

The response is a count and its variance is much larger than its mean. We
therefore use the negative binomial `NBI` family. It has two parameters:
`mu`, the conditional mean, and `sigma`, a dispersion parameter. Both use a
log link.

## A first model

The first model lets the mean motorcycle count vary smoothly over the year.
The cyclic cubic spline (`bs = "cc"`) joins smoothly at the beginning and end
of the annual cycle.

```{r}
m1 <- gamlss2(
  bikes ~ s(yday, bs = "cc", k = 20),
  data = HarzTraffic,
  family = NBI,
  trace = FALSE
)
```

Only one right-hand side is supplied, so it is used for the first family
parameter, `mu`. The remaining parameter, `sigma`, receives an intercept-only
predictor. More generally, the right-hand sides separated by `|` correspond to
the family parameters in their displayed order:

| Formula | Interpretation for `NBI` |
|---|---|
| `bikes ~ x` | model `mu` with `x`; keep `sigma` constant |
| `bikes ~ x | z` | model `mu` with `x` and `sigma` with `z` |
| `bikes ~ x | .` | use the predictor `x` for both parameters |

Thus, a model in which both the mean and dispersion vary seasonally can be
specified compactly as follows.

```{r}
m2 <- gamlss2(
  bikes ~ s(yday, bs = "cc", k = 20) | .,
  data = HarzTraffic,
  family = NBI,
  trace = FALSE
)
```

For families with more parameters, further right-hand sides can be appended.
For example, `y ~ x | . | . | .` repeats the predictor for all four parameters
of a four-parameter family. Omitting right-hand sides instead gives the
remaining parameters intercept-only predictors.

## Inspecting the fitted model

The model summary reports the family and links, linear coefficients, effective
degrees of freedom of the smooth terms, and model-fit information.

```{r}
summary(m2)
```

Calling `plot()` without further arguments displays the estimated effects for
all modeled distribution parameters.

```{r}
#| fig-height: 4
#| fig-width: 8
#| fig-align: center
par(mar = c(4, 4, 1, 1))
plot(m2)
```

Normalized quantile residuals should be approximately standard normal when the
conditional distribution is well calibrated. Because randomized quantile
residuals are used for discrete distributions, we set a seed to make the plots
reproducible.

```{r}
#| fig-height: 7
#| fig-width: 8
#| fig-align: center
set.seed(1328)
plot(m2, which = "resid")
```

These are in-sample diagnostics. They can reveal important model deficiencies,
but do not by themselves establish out-of-sample calibration.

## Distributional predictions

We now evaluate the fitted seasonal cycle on each day of a non-leap year. The
data encode day of year starting at zero, so plotting uses `yday + 1` for the
more familiar range from 1 to 365.

```{r}
nd <- data.frame(yday = 0:364)
```

By default, `predict()` returns the fitted distribution parameters on their
natural scales. Setting `drop = FALSE` keeps the result as a data frame (which is the default).

```{r}
parameters <- predict(m2, newdata = nd, drop = FALSE)
head(parameters)
```

For probabilities and other distributional summaries, `prodist()` returns one
[`distributions3`](https://cran.r-project.org/package=distributions3)
distribution object for each row of `newdata`.

```{r}
pf <- prodist(m2, newdata = nd)
```

For example, the fitted probability of more than 500 motorcycles on a day is
$1 - F(500)$.

```{r}
#| fig-height: 4
#| fig-width: 7
#| fig-align: center
prob_high <- 1 - cdf(pf, 500)

par(mar = c(4, 4, 1, 1))
plot(nd$yday + 1, prob_high, type = "l", lwd = 2,
  xlab = "Day of year",
  ylab = "Pr(Motorcycles > 500)")
```

Conditional quantiles can be computed directly from the fitted model. Here we
calculate the 5%, 50%, and 95% quantiles and add them to the observed counts.

```{r}
pq <- quantile(m2, newdata = nd, probs = c(0.05, 0.5, 0.95))
```

```{r}
#| fig-height: 4.5
#| fig-width: 6
#| fig-align: center
par(mar = c(4, 4, 1, 1))
plot(bikes ~ I(yday + 1), data = HarzTraffic,
  pch = 16, col = adjustcolor(1, 0.2),
  xlab = "Day of year", ylab = "Motorcycles")
matlines(nd$yday + 1, pq,
  lwd = 2, lty = c(3, 1, 2), col = 4)
legend("topright", c("5%", "50%", "95%"),
  lwd = 2, lty = c(3, 1, 2), col = 4, bty = "n")
```

These fitted distributions describe response variation conditional on the
estimated model. They do not include uncertainty from estimating coefficients,
smoothing parameters, or selecting the model.

## Relation to `gamlss`

`gamlss2` is a fresh implementation of the GAMLSS framework, but it can use the
family objects from `gamlss.dist`. Unlike the classic `gamlss()` workflow,
`gamlss2()` supports `mgcv` smooth terms directly and uses the `|` notation to
place the predictors for all distribution parameters in a single formula.

## Next steps

The [Family Objects](https://gamlss-dev.github.io/gamlss2/vignettes/families.html) 
vignette explains available and custom
response distributions. The
[Special Model Terms](https://gamlss-dev.github.io/gamlss2/vignettes/specials.html) 
vignette covers terms beyond standard
[`mgcv`](https://cran.r-project.org/package=mgcv)
smooths.

