---
title: "GAMLSS Modeling with Advanced Flexible Infrastructures"
author: "Mikis Stasinopoulos, Robert Rigby, Nikolaus Umlauf"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: gamlss2.bib
nocite: |
  @gamlss2:Rigby+Stasinopoulos:2005
vignette: >
  %\VignetteIndexEntry{First Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{gamlss2}
  %\VignetteKeywords{distributional regression, first steps}
  %\VignettePackage{gamlss2}
---

```{r preliminaries, echo=FALSE, message=FALSE, results="hide"}
library("gamlss2")
library("survival")
```

## Overview 

The primary purpose of this package is to facilitate the creation of advanced infrastructures
designed to enhance the GAMLSS modeling framework. Notably, the _gamlss2_ package represents a
significant overhaul of its predecessor, _gamlss_, with a key emphasis on improving estimation
speed and incorporating more flexible infrastructures. These enhancements enable the seamless
integration of various algorithms into GAMLSS, including gradient boosting, Bayesian estimation,
regression trees, and forests, fostering a more versatile and powerful modeling environment.

Moreover, the package expands its compatibility by supporting all model terms from the base
R _mgcv_ package. Additionally, the _gamlss2_ package introduces the capability to
accommodate more than four parameter families. Essentially, this means that users can now
specify any type of model using these new infrastructures, making the package highly
flexible and accommodating to a wide range of modeling requirements.

## Installation

The development version of _gamlss2_ is hosted on github at
<https://github.com/gamlss-dev/gamlss2>.
It can be installed via

```{r installation-github, eval=FALSE}
devtools::install_github("gamlss-dev/gamlss2")
```

## Introduction

This section gives a first quick overview of the functionality of the package and
demonstrates that the usual "look & feel" when using well-established model fitting 
functions like `glm()` is an elementary part of _gamlss2_, i.e., first steps and
basic handling of the package should be relatively simple. We illustrate the first steps
with _gamlss2_ using a data set taken from the _Regression Book_ [@Fahrmeir+Kneib+Lang+Marx:2013]
which is about prices of used VW Golf cars. The data is loaded with
```{r}
data("Golf", package = "bamlss")
head(Golf)
```
In this example the aim is to model the `price` in 1000 Euro. Using _gamlss2_ a first
linear model could be set up by first specifying a model formula
```{r}
f <- price ~ age + kilometer + tia + abs + sunroof
```
afterwards the model is estimated by
```{r}
library("gamlss2")
b1 <- gamlss2(f, family = NO, data = Golf)
```
The summary statistics can be shown with
```{r}
summary(b1)
```
For comparison, we estimate the same model using the _gamlss_ package.
```{r}
library("gamlss")
b2 <- gamlss(f, family = NO, data = Golf)
summary(b2)
```
Note that the estimated coefficients for both implementations, as well as standard errors
and p-values are almost the same. Similar for confidence intervals, we get
```{r}
confint(b1)
confint(b2)
```

## Munich Rent GAMLSS

To illustrate the functionality of the _gamlss2_ package, we use the _rent99_ dataset,
which is included in the _gamlss.data_ package.
```{r}
data("rent99", package = "gamlss.data")
```
We estimate models using the _GA_ distribution.
```{r}
## Model formula.
f <- rent ~ s(area) + s(yearc) + location + bath + kitchen |
   s(area) + s(yearc) + location + bath + kitchen

## Estimation.
start <- proc.time()
b1 <- gamlss2(f, data = rent99, family = GA)
proc.time() - start

## Estimation with gamlss.
start <- proc.time()
b2 <- gamlss(rent ~ pb(area) + pb(yearc) + location + bath + kitchen,
  sigma.formula =~ pb(area) + pb(yearc) + location + bath + kitchen,
  data = rent99, family = GA)
proc.time() - start

## Compare using the AIC.
AIC(b1, b2)
```
Note that _gamlss2_ is considerably faster in this example. Estimated effects
can be visualized with
```{r, out.width="100%", fig.width=12, fig.height=4}
plot(b1)
```
```{r, out.width="100%", fig.width=12, fig.height=8}
term.plot(b2, what = "mu", pages = 1, ask = FALSE)
```
```{r, out.width="100%", fig.width=12, fig.height=8}
term.plot(b2, what = "sigma", pages = 1, ask = FALSE)
```

## Big Data

To optimize computational efficiency when working with large datasets, we leverage the
observation that the number of distinct observations for a given covariate is typically
significantly smaller than the total number of observations in the complete dataset. As a strategy,
we calculate design matrices only using these unique observations and implement the
exceptionally efficient algorithms presented in
@Lang+Umlauf+Wechselberger+Harttgen+Kneib:2012. This option can be used by setting
the argument `binning = TRUE`. In many instances, it is feasible to decrease the number of
distinct observations through rounding, such as rounding to two digits. In the context of
large datasets, the approximation error associated with this rounding is typically negligible.

Below, we present an example utilizing precipitation data obtained from German meteorological
stations, provided by the Deutscher Wetterdienst (DWD, [https://www.dwd.de/](https://www.dwd.de/)).
In this illustration, we employ a censored normal distribution to model precipitation amounts,
which can be constructed using the _gamlss.cens_ package.
```{r}
library("gamlss.cens")
gen.cens(NO, type = "left")
```
The compiled data set is available at
[WeatherGermany.rds](https://fileshare.uibk.ac.at/f/795c0a9739de4e06b3a5/?dl=1)
and contains the following covariates
```{r}
d <- readRDS("WeatherGermany.rds")
print(dim(d))
print(head(d))
```
The model can be estimated with
```{r, eval = FALSE}
## Create indicator for Surv() response.
d$cens <- d$Pre > 0

## Model formula.
f <- ~ s(year) + s(alt,k=4) +
  ti(yday,bs="cc") + ti(lon,lat,bs="tp",d=2,k=30) +
  ti(yday,lon,lat,bs=c("cc","tp"),d=c(1,2),k=c(10,30))

f <- rep(list(f), 2)
f[[1]] <- update(f[[1]], Surv(sqrt(Pre), cens, type = "left") ~ .)

## Estimate model.
b <- gamlss2(f, data = d, family = NOlc,
  binning = TRUE, light = TRUE, flush = FALSE)
```
Please be aware that setting the argument `light = TRUE` results in the exclusion of all
design matrices, the model frame, and fitted values from the returned object.
This is done to minimize the object's size. Note that the model is quite complex, as it includes
a three dimensional seasonal varying spatial effect.

The summary statistics gives
```{r, echo=FALSE}
b <- readRDS("PreModel.rds")
d$cens <- d$Pre > 0
```
```{r}
summary(b)
```
Visualize estimated effects
```{r, out.width="100%", fig.width=12, fig.height=12}
plot(b, image = TRUE)
```

```{r, eval = FALSE, echo=FALSE}
d$Pre[d$Pre < 0] <- 0
d$cens <- d$Pre > 0

gen.cens(NO, type = "left")

f <- ~ s(year) + s(alt,k=4) +
  ti(yday,bs="cc") + ti(lon,lat,bs="tp",d=2,k=30) +
  ti(yday,lon,lat,bs=c("cc","tp"),d=c(1,2),k=c(10,30))

f <- rep(list(f), 2)
f[[1]] <- update(f[[1]], Surv(sqrt(Pre), cens, type = "left") ~ .)

if(!file.exists("PreModel.rds")) {
  b <- gamlss2(f, data = d, family = NOlc,
    binning = TRUE, light = TRUE, flush = FALSE)
  saveRDS(b, file = "PreModel.rds")
} else {
  b <- readRDS("PreModel.rds")
}
```

For spatial prediction of probabilities over Germany we can download the
[_ETOPO Global Relief Model_](https://www.ncei.noaa.gov/products/etopo-global-relief-model)

```{r, eval = FALSE, echo=FALSE}
library("sf")

## Download altitude data.
if(!file.exists("etopo.tif")) {
  download.file("https://www.ngdc.noaa.gov/mgg/global/relief/ETOPO2022/data/30s/30s_bed_elev_gtif/ETOPO_2022_v1_30s_N90W180_bed.tif", "etopo.tif")
} else {
  etopo <- readRDS("etopo.rds")
}


```

## Exchanging the Optimizer

The default optimizer function in _gamlss2_ is `RS()`. The function has arguments
```{r, eval = FALSE}
RS(x, y, specials, family, offsets, weights, xterms, sterms, control)
```
where `x` is the linear effects design matrix, `y` is the response (a vector or matrix), `specials`
is a named list of special model terms, e.g., as created by smooth constructor `s()`. Now `xterms`
and `sterms` are named lists, each list element represents one parameter of the distribution as
specified in the `family`. One list elements holds the variable names of linear effects or the term
labels of specials (smooth) model terms. This way, no duplicated design matrices need to be setup.

The return value of `RS()` is of class `"gamlss2"` for which summary, extractor and plotting functions
are provided.

Note that function `RS()`may be exchanged, either using `gamlss2_control()` or if an `$optimizer`
function is specified within the family. This way, one could also integrate full Bayesian MCMC
methods, or boosting algorithms etc. Only the methods for the return value need to be written new
for such functions.

## New Special Terms

In similar fashion new "special" model terms can be included for `RS()`and `GC()`. All is needed
is to write methods for
```{r, eval = FALSE}
special_fit(x, ...)
special_predict(x, ...)
```
where `x` is a model term object as supplied in the formula in the `gamlss2()` call.

Example using existing `smooth.construct()` infrastructures, new `pb2()` model term
```{r, eval = FALSE}
pb2 <- function(x, df = NULL, lambda = NULL, max.df = NULL, control = pb2.control(...), ...)
{
  m <- c(control$degree, control$order)
  k <- control$inter + m[1L] + 1L
  sx <- s(x, bs = "ps", m = m, k = k, sp = lambda)
  sx$control <- control
  sx$label <- gsub("s(", "pb2(", sx$label, fixed = TRUE)
  sx$orig.label <- sx$label
  sx$localML <- TRUE
  return(sx)
}
```

Or, completely new model term, e.g., conditional inference forest constructor.
```{r, eval = FALSE}
## Special conditional inference forest constructor.
cf <- function(formula, ...)
{
  stopifnot(requireNamespace("partykit"))
  st <- list()
  ctr <- list(...)
  ntree <- ctr$ntree
  if(is.null(ntree))
    ntree <- 100L
  ctr$ntree <- NULL
  st$control <- do.call(partykit::ctree_control, ctr)
  st$formula <- formula
  st$term <- all.vars(formula)
  st$label <- paste0("cf(", paste0(gsub(" ", "", as.character(formula)), collapse = ""), ")")
  st$data <- model.frame(formula)
  st$ntree <- ntree
  class(st) <- c("special", "cf")
  return(st)
}

## The fitting function for the backfitting algorithm.
special_fit.cf <- function(x, z, w, y, eta, j, family, control, ...)
{
  f <- update(x$formula, response_z ~ .)
  x$data$response_z <- z
  rval <- list(
    "model" = partykit::cforest(formula = f, data = x$data, weights = w,
      control = x$control, ntree = x$ntree)
  )
  rval$fitted.values <- predict(rval$model)
  rval$shift <- mean(rval$fitted.values)
  rval$fitted.values <- rval$fitted.values - rval$shift
  rval$edf <- x$ntree
  class(rval) <- "cf.fitted"
  return(rval)
}

## Predict method.
special_predict.cf.fitted <- function(x, data, se.fit = FALSE, ...)
{
  if(se.fit) {
    ct <- lapply(seq_along(x$model$nodes), function(i) partykit::as.constparty(
      partykit::party(x$model$nodes[[i]], data = x$model$data, terms = x$model$terms,
      fitted = data.frame(
        `(response)` = x$model$fitted[["(response)"]],
        `(weights)` = x$model$weights[[i]],
        check.names = FALSE))
    ))
    p <- sapply(ct, predict, newdata = data)
    p <- apply(p, 1, quantile, prob = c(0.05, 0.5, 0.95))
    p <- t(p)
    colnames(p) <- c("lower", "fit", "upper")
  } else {
    p <- predict(x$model, newdata = data, type = "response")
  }
  p <- p - x$shift
  return(p)
}
```
Similar, a neural network model term, see `?special_terms`.

## Notes

* Package _gamlss_ "special" model terms in _gamlss2_? Support is already there
  but limited, e.g., using `pb()` confidence intervals cannot be computed.
* Write wrappers like `pb2()`, use `smooth.construct()`?
* Goal is to write more optimizer/sampler functions.
* Project idea, "non-cyclic" algorithm for fast stepwise selection of linear or smooth terms?
* `predict()` with confidence/prediction intervals? `se.fit = TRUE` does not work using new data?
* How should we compute confidence/prediction intervals?
* Considering _gamlss_, extend the examples similar to _gamlss2_.
* Check the manual pages carefully, e.g., do we need references on every manual page?
* Include the `NO()` distribution in _gamlss_ (or hidden default `.NO()`) so we can write examples for each
  family in _gamlss.dist_?

## References

