---
title: "First Steps"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: gamlss2.bib
nocite: |
  @Rigby+Stasinopoulos:2005
vignette: >
  %\VignetteIndexEntry{First Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{gamlss2}
  %\VignetteKeywords{distributional regression, first steps}
  %\VignettePackage{gamlss2}
---

```{r preliminaries, echo=FALSE, message=FALSE, results="hide"}
library("gamlss2")
```

The package is designed to follow the workflow of well-established model fitting functions
like `lm()` or `glm()`, i.e., the step of estimating full distributional regression models
is actually not very difficult.

To illustrate the workflow using _gamlss2_, we analyze the `HarzTraffic` data, where we model
the number of motorcycles (response `bikes`) at Sonnenberg in the Harz region of Germany.
The data can be loaded with
```{r harzdata}
data("HarzTraffic")
head(HarzTraffic)
```
The data consists of seasonal time information (variable `yday`) along with a number of
environmental variables (e.g. mean daily `temp`erature). As a first model,
we estimate a linear regression model with normal errors (which is the default)
```{r model1}
b <- gamlss2(bikes ~ temp + rain + sunshine + wind, data = HarzTraffic)
summary(b)
```
Note that the summary output is very similar to `lm()` and `glm()` with the main difference
being that summary outputs are provided for all parameters of the distribution. In this case,
the model is estimated using the `NO` family of the _gamlss.dist_ package, a two-parameter
distribution with parameters `mu` and `sigma`.

### Residual Diagnostics

Since we estimated a simple linear model with Gaussian errors up to now, we are assuming that
the distribution of the response variable, the number of motorcycles (`bikes`), follows a
normal distribution with constant variance. However, this assumption may not always hold true,
especially when the response variable is count data, which often exhibits overdispersion
or non-constant variance.

To assess whether the normal distribution with constant variance is appropriate, we can
start by examining diagnostic plots.
```{r resids}
plot(b)
```

These plots help us visually inspect the residuals for any deviations from the assumptions
of normality and constant variance.

