---
title: "Special Model Terms"
format:
  html:
    html-math-method: mathjax
    toc: true
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Special Model Terms}
  %\VignetteEngine{quarto::html}
  %\VignetteDepends{gamlss2}
  %\VignetteKeywords{distributional regression, special model terms}
  %\VignettePackage{gamlss2}
---

Special model terms extend a `gamlss2` formula with a model component that is
fitted inside the backfitting algorithm. Ordinary linear terms and `mgcv`
smooths cover many models; specials are useful when the fitting method needs its
own estimator, prediction method, state, or tuning parameters.

```{r preliminaries, echo=FALSE, message=FALSE, results="hide"}
library("gamlss2")
```

## Built-in special terms

The package already provides several specials. For example, `lo()` fits a
weighted local-polynomial smoother, `tree()`, `ct()`, and `cf()` fit tree-based
terms, `n()` fits a neural-network term, and `lin()`, `re()`, `random()`, and
`gnet()` provide structured linear or grouped effects. The `mgcv` terms `s()`,
`te()`, `ti()`, and `t2()` are also supported.

```{r}
air <- subset(airquality, !is.na(Ozone))

m <- gamlss2(
  Ozone ~ lo(~ Temp, span = 0.7) + s(Wind) | s(Temp),
  data = air,
  family = GA,
  trace = FALSE
)

summary(m)
```

Specials can occur in any distributional parameter formula. They are refitted
from the current working response and weights at each backfitting iteration,
then centered before being added to the additive predictor. Centering keeps the
intercept and the special term identifiable.

## The special extension

A custom special consists of three pieces:

1. a constructor used in the formula;
2. a `special_fit()` method that fits the term to the current working response
   and weights; and
3. a `special_predict()` method for fitted values on new data.

The constructor should return a list of class `c("special", "user")` containing
the formula, variables, data, controls, and anything else needed by fitting and
prediction. The reserved `user` name is recognized by `fake_formula()`, so it
can be used without changing package internals. For a package-level special,
register its name in `fake_formula()` and use a dedicated class instead.

A fitting method receives at least `x`, `z`, and `w`. The framework can also
supply `y`, `eta`, `j`, `family`, and `control`; accepting `...` keeps a method
compatible with these additional arguments. The returned list must contain
numeric `fitted.values`. It should usually also contain:

- `edf`, the effective degrees of freedom;
- `model`, the fitted underlying model;
- `shift`, the centering constant; and
- optionally `transfer`, for state passed to the next backfitting iteration.

A prediction method receives the fitted object and `data`, and returns a numeric
vector. If it supports `se.fit = TRUE`, it should return a data frame with a
`fit` column and, optionally, interval columns.

## A custom weighted local smoother

The following example implements a small `user()` term around `stats::loess()`.
It is deliberately compact: a production term should validate its controls and
handle missing values and extrapolation explicitly.

```{r}
user <- function(formula, span = 0.75, degree = 2) {
  if(!inherits(formula, "formula"))
    formula <- as.formula(paste("~", deparse1(substitute(formula))))
  st <- list(
    formula = formula,
    data = model.frame(formula),
    control = list(span = span, degree = degree),
    term = all.vars(formula),
    label = paste0("user(", paste(all.vars(formula), collapse = "+"), ")")
  )
  class(st) <- c("special", "user")
  st
}

special_fit.user <- function(x, z, w, control, ...) {
  fit_formula <- update(x$formula, response_z ~ .)
  dat <- x$data
  dat$response_z <- z
  dat$weights_w <- w
  fit <- loess(
    fit_formula,
    data = dat,
    weights = weights_w,
    span = x$control$span,
    degree = x$control$degree
  )
  fitted <- as.numeric(predict(fit, newdata = dat))
  shift <- mean(fitted)
  out <- list(
    model = fit,
    fitted.values = fitted - shift,
    shift = shift,
    edf = fit$trace.hat
  )
  class(out) <- "user.fitted"
  out
}

special_predict.user.fitted <- function(x, data, se.fit = FALSE, ...) {
  fit <- as.numeric(predict(x$model, newdata = data)) - x$shift
  if(se.fit) return(data.frame(fit = fit))
  fit
}
```

The `user` constructor is available while the formula is parsed because it is
one of the reserved special names. The corresponding methods are found by S3
dispatch during fitting and prediction.

```{r}
m <- gamlss2(
  Ozone ~ user(~ Temp, span = 0.7) | s(Temp),
  data = air,
  family = GA,
  trace = FALSE
)

summary(m)
predict(m, newdata = air[1:5, , drop = FALSE])
```

## Practical guidance

Use an existing special when it already provides the required estimator and
prediction behavior. Write a custom special when the term has a distinct
fitting algorithm, needs custom state, or wraps an external modelling method.
Keep the constructor lightweight, store all information needed for prediction,
return centered fitted contributions, and test both fitting and `newdata`
predictions. The help topic `?special_terms` documents the method arguments and
return-value contract; `?fake_formula` documents special-name registration.

