---
title: "Survival Models"
format: 
  html:
    html-math-method: mathjax
    toc: true
    number-sections: true
bibliography: gamlss2.bib
nocite: |
  @Rigby+Stasinopoulos:2005
vignette: >
  %\VignetteIndexEntry{Survival Models}
  %\VignetteEngine{quarto::html}
  %\VignetteDepends{gamlss2}
  %\VignetteKeywords{distributional regression, inference, forecasting}
  %\VignettePackage{gamlss2}
---

In survival analysis, we aim to model the time until an event occurs, which is typically
represented as a random variable $T$. The survival function $S(t)$ indicates the probability
that the event has not occurred by time $t$ and is defined as
$$
S(t) = P(T > t) = 1 - F(t),
$$
where $F(t)$ is the cumulative distribution function (CDF) of $T$.

The hazard function $\lambda(t)$, which describes the instantaneous risk of the event occurring
at time $t$, is defined as
$$
\lambda(t) = \frac{f(t)}{S(t)},
$$
where $f(t)$ denotes the probability density function of $T$.

## Survival models in the GAMLSS framework

Within the GAMLSS framework, survival models are specified by selecting a parametric distribution
for the survival time $T$, characterized by a density
$$
f(t \mid \boldsymbol{\theta}),
$$
where $f(\cdot)\ represents a parametric density suitable for modeling survival times and
$\boldsymbol{\theta} = (\theta_1, \ldots, \theta_K)^\top$ is the vector of distributional
parameters. The choice of distribution implicitly determines the associated survival and hazard
functions.

To incorporate covariates, the distributional parameters are modeled as functions of explanatory
variables $\theta_k(\mathbf{x})$, $k = 1, \ldots, K$. Specifically, we use GAM-type predictors
of the form
$$
\eta_k(\mathbf{x}) = f_{1k}(\mathbf{x}) + \dots + f_{J_k k}(\mathbf{x}),
$$
which are linked to the parameters through suitable link functions,
$$
h_k\!\left(\theta_k(\mathbf{x})\right) = \eta_k(\mathbf{x}).
$$
The functions $f_{jk}(\cdot)$, $j = 1, \ldots, J_k$, may represent nonlinear smooth effects
(typically estimated using regression splines), linear effects, random effects, or other structured
components. This formulation allows for a flexible and interpretable modeling of how covariates
influence the full distribution of survival times.

Model estimation in GAMLSS is typically carried out using (penalized) maximum likelihood
estimation. For right-censored survival data with observations $i = 1, \ldots, n$, the
likelihood function is given by
$$
L(\boldsymbol{\beta}, \boldsymbol{\theta}) =
\prod_{i=1}^{n}
\left[ f(t_i \mid \boldsymbol{\theta}(\mathbf{x}_i)) \right]^{\delta_i}
\left[ S(t_i \mid \boldsymbol{\theta}(\mathbf{x}_i)) \right]^{1 - \delta_i},
$$
where $\delta_i$ is the censoring indicator for the $i$-th observation, taking the value
1 if the event is observed and 0 if the observation is right-censored. This likelihood formulation
naturally accommodates both observed and censored survival times and enables efficient estimation
of the model parameters.


## LeukSurv data example

In this example, we analyze the `LeukSurv` data provided in the `spBayesSurv` package.
This dataset contains survival information on patients diagnosed with leukemia, where the main
variables include `time` (the survival time), `cens` (the censoring indicator, with `1`
indicating the event has occurred and `0` indicating right-censoring), `age`, `sex`, `wbc`
(white blood cell count), `tpi` (tumor proliferative index), and spatial coordinates
`xcoord` and `ycoord` representing the location of the patients. These covariates provide both
demographic and clinical information that can be used to model the survival times and the
associated risks.

The dataset is particularly useful for illustrating spatial survival modeling, as it allows
us to incorporate spatial covariates (coordinates) into the survival model, capturing potential
geographical variations in survival probabilities. We will first demonstrate how to estimate
simple survival curves using `gamlss2`, followed by a full spatial survival model where
the geographical coordinates are explicitly modeled to investigate
any spatial effects on survival. 

We begin by loading the necessary packages and the `LeukSurv` dataset. These packages provide
functions for survival analysis, censored data handling, and spatial data manipulation, which are
essential for the analysis that follows. Specifically, `gamlss2` will be used for flexible
survival model estimation, while `gamlss.cens` manages the censored data structure.
Additionally, `sf` and `stars` will facilitate spatial data processing and visualization.

```{r packages}
#| message: false
#| results: hide
#| echo: true
## install required packages
pkg <- c("spBayesSurv", "gamlss.cens", "sf", "stars")
for(p in pkg) {
  if(!(p %in% installed.packages())) install.packages(p)
}
## load required packages
library("gamlss2")
library("gamlss.cens")
library("sf")
library("stars")

## load the LeukSurv data
data("LeukSurv", package = "spBayesSurv")
```

## Generating a censored family

To handle censored data in the `gamlss2` framework, we first need to generate a censored
family based on a parametric distribution. In this case, we use the Weibull distribution
(denoted as `WEI` in `gamlss.dist`). This involves transforming the standard Weibull
family into one that can handle censored observations.

We accomplish this by using the `gen.cens()` function provided in the `gamlss.cens` package,
which adapts the distribution to accommodate censoring. Then, we create a censored family
using the `cens()` function.

```{r}
gen.cens(WEI)
fam <- cens(WEI)
```

## Estimating time-only models

To begin, we estimate a simple survival model without covariates (an intercept-only model).
This is done both with `gamlss2` for flexible distributional modeling, and with the more
traditional `survfit()` function from the `survival` package for comparison.

In `gamlss2`, the model is specified using the `Surv(time, cens)` function, where `time`
is the survival time, and `cens` is the censoring indicator. We fit the model assuming a
Weibull distribution for survival times.

```{r}
## using gamlss2
m1 <- gamlss2(Surv(time, cens) ~ 1, family = fam, data = LeukSurv)

## and now with survfit
s1 <- survfit(Surv(time, cens) ~ 1, data = LeukSurv)
```

## Predicting survival curves

Once the models are fitted, we can predict survival probabilities based on the estimated parameters.
In `gamlss2`, the predicted parameters from the model are used to compute survival probabilities.
We then compare these with the survival probabilities obtained from the Kaplan-Meier estimate
using `survfit()`.

In the case of `gamlss2`, we extract the predicted parameters and compute the
survival probabilities with

```{r}
## predict parameters from the gamlss2 model
par <- predict(m1, newdata = LeukSurv)

## compute survival probabilities
p1 <- 1 - family(m1)$cdf(Surv(LeukSurv$time, rep(1, nrow(LeukSurv))), par)
```

For the `survfit()` model, we use the step function to extract the survival
probabilities at each time point.

```{r}
## step function for Kaplan-Meier survival estimates
f <- stepfun(s1$time, c(1, s1$surv))
p2 <- f(LeukSurv$time)
```

## Plotting the estimated survival curves

Finally, we plot the estimated survival curves from both models to visually compare the results.
This will allow us to see how the flexible parametric model from `gamlss2` compares with the
non-parametric Kaplan-Meier estimate.

```{r}
matplot(LeukSurv$time, cbind(p1, p2),
  type = "l", lty = 1, lwd = 3, col = c(1, 4),
  xlab = "Time", ylab = "Survival Prob(t > Time)",
  main = "Estimated Survival Curves")
legend("topright", c("gamlss2", "survfit"),
  lwd = 3, col = c(1, 4), bty = "n")
```

In this plot, we observe that the estimated survival curves are generally similar. However,
the Kaplan-Meier estimate shows a slightly steeper decline in survival probabilities early on
and flattens out more towards the end. This difference likely reflects the nature of the
Kaplan-Meier estimator, which is non-parametric and purely empirical, capturing the observed
data trends without any underlying assumptions about the distribution of survival times.
In contrast, the parametric model from `gamlss2` provides a more structured approach, allowing
for smoother, distribution-driven survival estimates that incorporate the shape of the
underlying Weibull distribution.

## Spatial survival model

Next, we extend our analysis by estimating a full spatial survival model. This model
incorporates both clinical covariates (such as `age`, `sex`, `wbc`, and `tpi`) and spatial
information represented by the coordinates (`xcoord` and `ycoord`). The inclusion of spatial
covariates allows us to assess potential geographic variation in survival probabilities,
which could reveal spatial clustering or regional effects on survival outcomes.

The model is specified using the formula below, where the response is the censored survival
time (`Surv(time, cens)`), and smooth terms (`s()`) are used to model nonlinear effects of
continuous covariates like `age`, `wbc`, `tpi`, and the spatial coordinates. The pipe
symbol (`|`) separates the formulas for the parameters of the distribution,
allowing for flexible covariate effects on survival times.

```{r}
## model formula
f <- Surv(time, cens) ~ sex + s(age) + s(wbc) + s(tpi) + s(xcoord, ycoord) | .

## estimate the spatial survival model
b2 <- gamlss2(f, family = fam, data = LeukSurv)
```
## Visualizing effects & residuals

To explore the impact of the covariates and spatial coordinates, we can plot the
estimated effects from the model. This visualization helps us understand how each covariate,
including the spatial effect, influences survival outcomes.

```{r}
#| fig-height: 7.5
#| fig-width: 7
#| fig-align: center
par(mar = c(4, 4, 3, 1))
plot(b2)
```

To assess the model fit, we generate diagnostic plots of the quantile residuals.
These plots allow us to evaluate the adequacy of the model and check for any potential issues.

```{r}
#| fig-height: 6
#| fig-width: 6
#| fig-align: center
plot(b2, which = "resid")
```

The plot indicates that the Weibull distribution provides a reasonable starting point for modeling
the survival data. However, the wormplot reveals some departures from the assumed distribution,
suggesting that alternative specifications should be considered to achieve a better overall fit.
Within the GAMLSS framework, a wide range of parametric distributions such as the log-normal,
gamma, or generalized gamma, can be readily evaluated. Exploring these alternatives is
particularly advisable when residual diagnostics indicate systematic deviations or potential
misspecification of the underlying hazard function.

## Visualize the spatial effect.

To visualize the estimated spatial effect, we construct a map of the New England region using
polygon boundary data from the `spBayesSurv` package.

```{r}
#| fig-align: center
## read the map of new west england.
file <- system.file("otherdata/nwengland.bnd", package = "spBayesSurv")
d <- readLines(file)

## transform the polygons to a list().
id <- grep('\"', d, fixed = TRUE)
polys <- list()
for(i in 1:length(id)) {
  j <- strsplit(d[id[i]], ",")[[1]][2]
  if(i < length(id))
    polys[[j]] <- d[(id[i] + 1):(id[i + 1] - 1)]
  else
    polys[[j]] <- d[(id[i] + 1):length(d)]
}

polys <- lapply(polys, function(x) {
  tf <- tempfile()
  writeLines(x, tf)
  pol <- as.matrix(read.csv(tf, header = FALSE))
  unlink(tf)
  return(st_polygon(list(pol)))
})

## transform to sf object
polys_sfc <- st_sfc(polys)
map <- st_sf(geometry = polys_sfc)
map$id <- names(polys)
map$district <- 1:nrow(map)

## plot the map
par(mar = rep(0, 4))
plot(st_geometry(map))
```

A regular grid of locations is then sampled within the study area, and model predictions are
obtained at these locations while holding all non-spatial covariates fixed at representative values.

Based on the predicted distributional parameters, survival probabilities at 365 days are
computed and displayed as a raster map overlaid with the district boundaries. This visualization
highlights spatial variation in survival that is captured by the model, facilitating an intuitive
assessment of geographical heterogeneity.

```{r}
#| fig-align: center
## sample coordinates for plotting
co <- st_sample(map, size = 10000, type = "regular")

## create new data for prediction
nd <- as.data.frame(st_coordinates(co))
names(nd) <- c("xcoord", "ycoord")
nd$sex <- 1
nd$wbc <- mean(LeukSurv$wbc)
nd$tpi <- mean(LeukSurv$tpi)
nd$age <- mean(LeukSurv$age)

## predict parameters
par <- predict(b2, newdata = nd)

## compute survival probabilities
p2 <- 1 - family(b2)$cdf(Surv(rep(365, nrow(nd)), rep(1, nrow(nd))), par)

## plot on map
sp <- st_sf(geometry = co)
sp$survprob <- p2

## reference system, only used for plotting
sp <- st_set_crs(sp, 4326)
map <- st_set_crs(map, 4326)

## plot as raster map
spr <- st_as_stars(sp)

par(mar = c(0, 0, 5, 0))
plot(spr, main = "Survival Probability P(T > 365)",
  reset = FALSE, nbreaks = 100)
plot(st_geometry(map), add = TRUE)
box()
```

